Detection and characterisation of stellar multiplicity with \Gaia

### 2018-06-05 article/ms.tex
whoa

\title{Detection and characterisation of stellar multiplicity from \Gaia\ data}

@@

\begin{abstract}
Most stars are in binary systems. The occurrence rate and properties of stellar 
multiples (e.g., binaries, triples) is relevant to many fields of astrophysics, 
but the effects of stellar multiplicity are commonly ignored in the analysis of 
most stellar and galactic surveys. Here we use \Gaia\ data to identify 
\todo{\NumberOfHigherOrderSystems} stellar multiple systems from radial velocity
and astrometric jitter, photometry, and intrinsic luminosity. Our sample includes
\todo{\NumberOfSingleLinedSystems} unresolved binary systems detected solely
from radial velocity or astrometric jitter, which have orbital periods up to 
$\approx$3.5\,yrs. We show that the ratio of jitter in radial velocity and
astrometry can be calibrated to estimate inclination angles (and companion masses)
for thousands of systems. \todo{We comment on stellar multiplicity with stellar
metallicity}. \todo{We comment on systems with stellar remnants (e.g. NS/BHs).}
\end{abstract}

@@

In principle for well-sampled circular orbits this model would allow us to
identify binary star systems with orbital periods up to twice the observing
span in the second \Gaia\ data release ($\approx$21 months observing span; 
$\approx$44 month orbital periods). For higher-order star systems with long 
periods, or with mass ratios that result in a small velocity semi-amplitude $K$, 
under our assumptions these systems may be classified as single-star systems 
because the intrinsic radial velocity variation is within the expected 
variations for single-stars of a similar apparent magnitude and colour. We 
revisit this issue in Section \ref{sec:sb_limits}, where we explicitly define 
the probability that a \Gaia\ source is a higher-order star system $p(y|\textrm{SB1})$ 
within defined limits of orbital period (and other orbital parameters). Outside of this
parameter range, we are insensitive to distinguishing single-star systems from
higher-order SB1-type systems.

\begin{itemize}
    \item show jitter compared to RP flux, etc
    \item how the model was fit
    \item show results of the fit
    \item what things did we \emph{not} include in the model (eg difference between temperature and the temperature of the template used), and why
    \item results
\end{itemize}

@@

We calculated the completeness of radial velocity measurements (e.g., the
fraction of sources with reported radial velocities) as a function of all 
available source properties that might affect whether the radial velocity may
be reported or not. This included position ($\alpha$, $\delta$, $l$, $b$),
parallax, proper motions, apparent magnitudes, colours, properties of the radial
velocity templates, stellar parameters ($\teff$, $\radius$, $\luminosity$),
and other properties. The full list of properties investigated is described in
Appendix \ref{appendix:sb2}. In Figure \ref{fig:radial_velocity_completeness} 
we show the completeness as a function of some pertinent properties, which demonstrate 
that the radial velocity completeness is relatively flat until a star becomes 
too faint (low \rp\ flux), or is either too blue or too red. We adopt 
conservative limits for when the radial velocity completeness starts to drop 
with these properties, and we assume that any source within the following
parameter range

\begin{eqnarray}
    \mathrm{phot\,\,rp\,\,mean\,\,flux} & \in & [10^{4}, 10^8] \nonumber \\
    \mathrm{bp\,\,-rp} & \in & [X, Y] 
    \label{eq:sb2_criteria}
\end{eqnarray}
\noindent{}is likely to be a double-lined spectroscopic binary (SB2) if no 
radial velocity is reported. That is to say that we are explicitly assuming
that if a point source meets the criteria in Equation \ref{eq:sb2_criteria} 
and does not have a radial velocity measurement, then the point source is
an unresolved double-lined spectroscopic binary. In principle we could
make more realistic attempts to model the radial velocity completeness as
a function of stellar properties rather than simply stating ``sources within
this parameter range should have radial velocities unless they are SB2 systems'',
but the radial velocity completeness within our specified range is reasonably
flat.

@@

\subsection{Unresolved equal mass binaries: photometry}

@@

It is a pleasure to thank
    Adrian Price-Whelan,
	David W. Hogg,
	Lorenzo Spina, 
		and 
	Kevin C. Schlaufman.
This work was partly supported through the Australian Research Council 
through Discovery Grant DP160100637.
This work has made use of data from the European Space Agency (ESA) mission {\it
Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia} Data
Processing and Analysis Consortium (DPAC,
\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding for the DPAC
has been provided by national institutions, in particular the institutions
participating in the {\it Gaia} Multilateral Agreement.  This research was
started at the NYC Gaia DR2 Workshop at the Center for Computational
Astrophysics of the Flatiron Institute in 2018 April.

>>>>

\title{Detection and characterisation of stellar multiplicity with \Gaia}

@@

\begin{abstract}
Most stars have stellar companions. The occurrence rate and properties of stellar 
multiples (e.g., binaries, triples) is relevant to many fields of astrophysics, 
but the effects of stellar multiplicity are commonly ignored in the analysis of 
most stellar and galactic surveys. Here we use \Gaia\ data to identify 
\todo{\NumberOfHigherOrderSystems} stellar multiple systems from radial velocity
and astrometric jitter, photometry, and intrinsic luminosity. Our sample includes
\todo{\NumberOfSingleLinedSystems} unresolved binary systems detected solely
from radial velocity or astrometric jitter, which have orbital periods up to 
$\approx$3.5\,yrs. We show that the ratio of jitter in radial velocity and
astrometry can be calibrated to estimate inclination angles (and companion masses)
for thousands of systems. \todo{We comment on stellar multiplicity with stellar
metallicity}. \todo{We comment on systems with stellar remnants (e.g. NS/BHs).}
\end{abstract}

@@

In principle for well-sampled circular orbits this model would allow us to
identify binary star systems with orbital periods at least up to twice the observing
span in the second \Gaia\ data release ($\approx$21 months observing span; 
allowing for well-characterised detections for systems with orbital periods up to $\approx$44 months).

For higher-order star systems with longer periods, we can still find that a
stellar multiple system is more likely than a single star scenario, but our
estimate of the orbital properties (e.g., the semi-amplitude $K$) will be biased
to lower values because we have not fully sampled half of the orbital period.
At even longer orbital periods, or for stellar multiples with mass ratios that
produce a small velocity semi-amplitude $K$, under our assumptions these systems 
may be classified as single-star systems because the intrinsic radial velocity variation is within the expected 
variations for single stars of a similar apparent magnitude and colour. We 
revisit this issue in Section \ref{sec:sb_limits}, where we explicitly define 
the probability that a \Gaia\ source is a SB1-type system $p(\textrm{SB1}|y)$ 
within defined limits of orbital period (and other orbital and source parameters). 
Outside of this parameter range, we are insensitive to distinguishing single-star 
systems from higher-order SB1-type systems.

% Model fitting.

% Figure: fitting the radial velocity variance as a function of stellar properties.

% Figures: Binarity as a function of source properties.

% Figure: h-r diagram of SB1 binary fraction

% Figure: histogram of binaries as a fucntion of the Rv excess

@@

We calculated the completeness of radial velocity measurements (e.g., the
fraction of sources with reported radial velocities) as a function of all 
available source properties that might affect whether the radial velocity may
be reported or not. This included position ($\alpha$, $\delta$, $l$, $b$),
parallax, proper motions, apparent magnitudes, colours, properties of the radial
velocity templates, stellar parameters ($\teff$, $\radius$, $\luminosity$),
and other properties. The full list of properties investigated is described in
Appendix \ref{appendix:sb2}. In Figure \ref{fig:radial_velocity_completeness} 
we show the completeness as a function of some pertinent properties, which demonstrate 
that the radial velocity completeness is relatively flat until a star becomes 
too faint (low \rp\ flux), or is either too blue or too red. We adopt 
conservative limits for when the radial velocity completeness starts to drop 
with these properties, and we assume that any source within the following
parameter range
\begin{eqnarray}
    \mathrm{phot\,\,rp\,\,mean\,\,flux} & \in & [10^{4}, 10^8] \nonumber \\
    \mathrm{bp-rp} & \in & [X, Y] 
    \label{eq:sb2_criteria}
\end{eqnarray}
\noindent{}is likely to be a double-lined spectroscopic binary (SB2) if no 
radial velocity is reported. That is to say that we are explicitly assuming
that if a point source meets the criteria in Equation \ref{eq:sb2_criteria} 
and does not have a radial velocity measurement, then the point source is
an unresolved double-lined spectroscopic binary. In principle we could
make more realistic attempts to model the radial velocity completeness as
a function of stellar properties rather than simply stating ``sources within
this parameter range should have radial velocities unless they are SB2 systems'',
but the radial velocity completeness within our specified range is reasonably
flat.

@@

\subsection{Unresolved near-equal mass binaries: photometry}
\label{sec:pb_methods}

If a binary system is observed nearly face-on (at an inclination angle 
$i \approx 0^\circ$) then there is will be no detectable excess radial 
velocity variations. In principle there may be detectable astrometric
variations, but most near-equal mass binaries would be detected more 
reliably through intrinsic magnitudes that are anomalously lower (brighter)
and bluer colours than what would be expected for a single star system.

\todo{How the fuck am I going to model this, exactly?}

@@

It is a pleasure to thank
    Adrian Price-Whelan (Princeton University),
	Lorenzo Spina (Monash University), 
		and 
	Kevin C. Schlaufman (Johns Hopkins University).
This work was partly supported through the Australian Research Council 
through Discovery Grant DP160100637.
This work has made use of data from the European Space Agency (ESA) mission {\it
Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia} Data
Processing and Analysis Consortium (DPAC,
\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding for the DPAC
has been provided by national institutions, in particular the institutions
participating in the {\it Gaia} Multilateral Agreement.  This research was
started at the NYC Gaia DR2 Workshop at the Center for Computational
Astrophysics of the Flatiron Institute in 2018 April.
